---
title: "Cleaning Jeopardy Data"
author: "Stephen Kaluzny"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

## Setup

```{r dplyr}
library("dplyr", quietly= TRUE, warn.conflicts = FALSE)
```

## Data

Read the delimited data file.
```{r read_data}
dfile <- "game_player_score.txt"
dpath <- here::here(file.path("data", dfile))
d <- read.table(dpath, header=TRUE, sep="|", quote='"', comment="")
d[["date"]] <- as.Date(d[["date"]])
```

The data has `r nrow(d)` observations and `r ncol(d)` variables.

There are `r length(unique(d$game))` unique games
and `r length(unique(d$show))` unique shows.

The variable names and type.
```{r data_type}
sapply(d, class)
```


## Data Clean Up

The `occupation`, `from`, and `last` variables need work.
We extract them, do the fix up and then put them back into the data.frame.

```{r extract_from_occupation_last}
from2 <- d[["from"]]
occupation2 <- d[["occupation"]]
last2 <- d[["last"]]
```

The scraped data has errors with the `last` name suffix "JR." being included
with `occupation`

```{r clean01}
# Fix occupation and from
# occupation includes "JR.," at start, add to end of last:
indx_jr <- grep("\\bjr\\., ", occupation2, ignore.case=TRUE)
if(length(indx_jr) > 0) {
  occupation2[indx_jr] <- gsub("^JR\\., ", "", occupation2[indx_jr])
  last2[indx_jr] <- paste0(last2[indx_jr] , ", Jr.")
}
```

The word `Originally` gets mixed in with `occupation`.
Just remove it.
The word `From` ends up in `occupation`, remove it.

```{r clean02}
# occupation include "Originally" at the end, remove:
occupation2 <- gsub(" Originally$", "", occupation2)
# No occupation, just the word "From", set to NA:
indxFrom <- which(occupation2 == "From")
if(length(indxFrom) > 0) {
  occupation2[indxFrom] <- NA
}
# No occupation, just the word "Originally", set to NA:
indxOriginally <- which(occupation2 == "Originally")
if(length(indxOriginally) > 0) {
  occupation2[indxOriginally] <- NA
}
# occupation starts with "From" and from is NA, swap:
indxOccHasFromStart <- grep("^From ", occupation2)
indxFromNA <- which(is.na(from2))
indxOccToFrom <- intersect(indxOccHasFromStart, indxFromNA)
if(length(indxOccToFrom) > 0) {
  from2[indxOccToFrom] <- occupation2[indxOccToFrom]
  occupation2[indxOccToFrom] <- NA
}
# occupation ends with "From", remove it:
indxOccHasFromEnd <- grep("\\sFrom$", occupation2)
if(length(indxOccHasFromEnd) > 0) {
  occupation2[indxOccHasFromEnd] <- gsub("\\sFrom$", "",
    occupation2[indxOccHasFromEnd])
}
# Remove "From" from the from variable:
from2 <- gsub("^From ", "", from2)
from2 <- gsub(" From$", "", from2)
```
From previous working with the J Archive data some misspelled
occupation and location values.
Correct with the `fixfrom` function.

```{r clean03}
# Fix known misspelled / problem from values:
fixfrom <- function(x, old, new) {
  indx <- which(x == old)
  if(length(indx) > 0) {
    x[indx] <- new
  }
  x
}
from2 <- fixfrom(from2, "New York City", "New York, New York")
from2 <- fixfrom(from2, "New York City, New York", "New York, New York")
from2 <- fixfrom(from2, "The Bronx", "New York, New York")
from2 <- fixfrom(from2, "The Bronx, New York", "New York, New York")
from2 <- fixfrom(from2, "Brooklyn, New York", "New York, New York")
from2 <- fixfrom(from2, "Far Rockaway, New York", "New York, New York")
from2 <- fixfrom(from2, "Forest Hills, New York", "New York, New York")
from2 <- fixfrom(from2, "Flushing, New York", "New York, New York")
from2 <- fixfrom(from2, "Fresh Meadows, New York", "New York, New York")
from2 <- fixfrom(from2, "Harlem, New York", "New York, New York")
from2 <- fixfrom(from2, "Jackson Heights, New York", "New York, New York")
from2 <- fixfrom(from2, "Jamaica, New York", "New York, New York")
from2 <- fixfrom(from2, "Kew Gardens, New York", "New York, New York")
from2 <- fixfrom(from2, "Little Neck, New York", "New York, New York")
from2 <- fixfrom(from2, "Rockaway Park, New York", "New York, New York")
from2 <- fixfrom(from2, "Rockaway Beach, New York", "New York, New York")
from2 <- fixfrom(from2, "Manhattan, New York", "New York, New York")
from2 <- fixfrom(from2, "Queens, New York", "New York, New York")
from2 <- fixfrom(from2, "Bronx, New York", "New York, New York")
from2 <- fixfrom(from2, "Staten Island, New York", "New York, New York")
from2 <- fixfrom(from2, "Downers Grove, Il", "Downers Grove, Illinois")
from2 <- fixfrom(from2, "Saint-Laurent, Montreal", "Montreal, Canada")
from2 <- fixfrom(from2, "West Hartford", "West Hartford, Connecticut")
from2 <- fixfrom(from2, "Lynn Massachusetts", "Lynn, Massachusetts")
from2 <- fixfrom(from2, "San Francisco", "San Francisco, California")
from2 <- fixfrom(from2, "San Jose California", "San Jose, California")
from2 <- fixfrom(from2, "Kansas City", "Kansas City, Missouri")
from2 <- fixfrom(from2, "New York State", "New York")
from2 <- fixfrom(from2, "Carmel Valley", "Carmel Valley, California")
from2 <- fixfrom(from2, "Tampa", "Tampa, Florida")
from2 <- fixfrom(from2, "St. Louis", "St. Louis, Missouri")
from2 <- fixfrom(from2, "ST. Louis", "St. Louis, Missouri")
from2 <- fixfrom(from2, "Easton Maryland", "Easton, Maryland")
from2 <- fixfrom(from2, "Nellis Air Force Base, Las Vegas", "Las Vegas, Nevada")
from2 <- fixfrom(from2, "Yellowstone National Park, Wyoming And Montana", "Yellowstone National Park, Wyoming")
from2 <- fixfrom(from2, "Rowland's Castle, Hampshire, England",
  "Rowland's Castle, England")
from2 <- fixfrom(from2, "Leeds, Yorkshire England", "Leeds, England")
from2 <- fixfrom(from2, "Sault StE. Marie, Canada", "Sault Ste. Marie, Canada")
from2 <- fixfrom(from2, "Surrey, British Colombia", "Surrey, British Columbia")
from2 <- fixfrom(from2, "Seattle Washington", "Seattle, Washington")
from2 <- fixfrom(from2, "Cassopolis, Mighigan", "Cassopolis, Michigan")
from2 <- fixfrom(from2, "Trinidad And Tobago And Now In New York",
  "New York, New York")
from2 <- fixfrom(from2, "Jamaica, Now Living In Toronto, Ontario, Canada",
  "Toronto, Canada")
from2 <- fixfrom(from2, "Sandy Mannix", NA)
```

For players who are now stationed somewhere (military), use the original
as the `from` value.
```{r now_stationed}
from2 <- sub(", Now Stationed At.*", "", from2)
from2 <- sub(", Now Stationed In.*", "", from2)
```

Drop province from Canadian players.
If Canada is not included, add it after the city.

```{r canada, eval=TRUE}
from2 <- sub(" Ontario Canada", " Ontario, Canada", from2)
for(p in c("Alberta", "British Columbia", "Manitoba", "New Brunswick", 
  "Newfoundland And Labrador", "Nova Scotia", "Ontario",
  "Prince Edward Island", "Quebec", "Saskatchewan", "Yukon Territory")) {
  from2 <- sub(paste0(", ", p, ", Canada"), ", Canada", from2)
  from2 <- sub(paste0(", ", p), ", Canada", from2)
}
```

Fix city Saint abbreviation from `ST.` to `St.`
```{r clean04}
from2 <- sub("ST. ", "St. ", from2, fixed=TRUE)
```

Now use the revised `last2`, `from2` and `occupation2`
to replace `last`, `from` and `occupation` (resp.) in the original data.frame.

```{r update_data.frame}
d[["last"]] <- last2
d[["from"]] <- from2
d[["occupation"]] <- occupation2
```

## Name
Create a `name` variable by pasting together `first` and `last`,
add it to the data
and from `first` and `last`.

```{r name}
d[["first"]] <- gsub("\\\\", "", d[["first"]])
d[["name"]] <- paste(d[["first"]], d[["last"]])
d <- select(d, -first, -last)
```

## City / State
To analyze data at the state level,
split `from` into `city` and `state`.
For players outside the Unites States, `state` will contain their country.

```{r citystate}
citystate <- stringr::str_split(d[["from"]], pattern=",", simplify=TRUE)
city <- citystate[,1]
state <- trimws(citystate[,2])
indx_no_city <- which(state == "")
state[indx_no_city] <- city[indx_no_city]
city[indx_no_city] <- NA
d[["city"]] <- city
d[["state"]] <- state
d <- select(d, -from)
```

## Non-Regular Jeopardy Games

Over the years Jeopardy has had a number of special games.
These include an annual Tournament of Champions,
Celebrity Jeopardy, College and Children Tournament.
These games do not follow the standard winner moves on format.
As such, they will be removed.

To find these non-regular games the comments variable is searched
for particular words used in describing these games.

```{r find_non_regular}
indx_tournament <- grep("tournament", d[["comments"]], ignore.case=TRUE)
indx_championship <- grep("championship", d[["comments"]], ignore.case=TRUE)
indx_super_jeopardy <- grep("Super Jeopardy", d[["comments"]], ignore.case=TRUE)
indx_celebrity <- grep("celebrity", d[["comments"]], ignore.case=TRUE)
indx_battle <- grep("\\bbattle\\b", d[["comments"]], ignore.case=TRUE)
indx_kids <- grep("\\bkids week\\b", d[["comments"]], ignore.case=TRUE)
indx_power <- grep("\\bpower players\\b", d[["comments"]], ignore.case=TRUE)
indx_backtoschool <- grep("\\bback to school\\b", d[["comments"]], ignore.case=TRUE)
indx_allstar <- grep("All-Star", d[["comments"]], ignore.case=TRUE)
indx_goat <- grep("\\bgreatest of all time\\b", d[["comments"]], ignore.case=TRUE)
indx_quarterfinal <- grep("quarterfinal", d[["comments"]], ignore.case=TRUE)
indx_semifinal <- grep("semifinal", d[["comments"]], ignore.case=TRUE)
indx_final <- grep("\\bfinal game", d[["comments"]], ignore.case=TRUE)
```
The IBM challenge was a series of games where contestants played against
IBM Watson computer.

```{r watson_games}
indx_IBM_challenge <- c(
  which(d[["game"]] == 3575),
  which(d[["game"]] == 3576),
  which(d[["game"]] == 3577))
```

We combine all these sesarch results (indices) and take the union.
```{r indx_not_regular}
indx_not_regular <- sort(unique(c(
  indx_tournament,
  indx_championship,
  indx_super_jeopardy,
  indx_celebrity,
  indx_battle,
  indx_kids,
  indx_power,
  indx_backtoschool,
  indx_allstar,
  indx_goat,
  indx_IBM_challenge,
  indx_quarterfinal,
  indx_semifinal,
  indx_final
)))
```

There are some regular games where the words Tournament or Battle were
used in the comments.
We want to keep those games.

```{r keep_games}
# Keep games: 3970, 4085 4305, 4364
# Regular game with the word Tournament or Battle in comment:
indx_3970 <- which(d[["game"]] == 3970)
indx_not_regular <- setdiff(indx_not_regular, indx_3970)
indx_4085 <- which(d[["game"]] == 4085)
indx_not_regular <- setdiff(indx_not_regular, indx_4085)
indx_4305 <- which(d[["game"]] == 4305)
indx_not_regular <- setdiff(indx_not_regular, indx_4305)
indx_4364 <- which(d[["game"]] == 4364)
indx_not_regular <- setdiff(indx_not_regular, indx_4364)
```

Early Trebek pilots that duplicate show numbers 1 and 2, should be dropped.
```{r drop games}
indx_1309 <- which(d[["game"]] == 1309)
indx_1317 <- which(d[["game"]] == 1317)
indx_not_regular <- c(indx_not_regular, indx_1309, indx_1317)
game_not_regular <- d[indx_not_regular, ][["game"]]
saveRDS(game_not_regular, "/tmp/game_not_regular.rds")
```

Now remove all the non-regular games.
Use the regular games data set going forward.

```{r drop_non_regular}
d <- d[-indx_not_regular,]
```

## Additional Variables

### Season

Add a season variable.
The regular season starts in September
and typically ends in July of the following year.

```{r season}
startyear <- lubridate::year(min(d[["date"]]))
endyear <- lubridate::year(max(d[["date"]]))
endmonth <- lubridate::month(max(d[["date"]]))
if(endmonth > 8) endyear <- endyear + 1
season_breaks <- seq(from=as.Date(paste0(startyear, "-09-01")),
  to=as.Date(paste0(endyear, "-09-01")),
  by="year")
season <- cut(d[["date"]], breaks=season_breaks,
  labels=seq_len(length(season_breaks) - 1))
d[["season"]] <- as.numeric(season)
```

Check start and end games for each season.

```{r season_min_max}
d |> group_by(season) |>
  summarise(start=min(date), end=max(date)) |>
  as.data.frame()
```

### Player Id
To facilitate analysis of on player basis,
no matter how many games they played,
we will create a player id by pasting together the
`name`, `occupation`, `city` and `state` variables.

```{r playerid}
d[["playerid"]] <- with(d,
  paste(name, occupation, city, state, sep="."))
```

Analysis of `playerid` indicated that it uniquely identified players
e.g. show numbers and dates were grouped together.

Using the `playerid` we can create the number of regular games played
for each player (`ngames`).

```{r ngames}
d <- d |> group_by(playerid) |>
  mutate(ngames = n()) |>
  ungroup()
```
  
### iswinner

Add a logical variable, `iswinner`, to denote the winner of game.
```{r winner}
isWinner <-
function (x)
{
    r <- rank(x)
    if (all(r == 2)) {
        r <- c(4, 4, 4)
    }
    (r == 3) | (r == 2.5) | (r == 4)
}
d <- d %>%
  group_by(game) %>%
  mutate(iswinner = isWinner(final)) %>%
  ungroup()
```

Two games (174 and 5167) resulted in no winner,
all players had 0 for their final score.
Set `iswinner` to `FALSE` for all players in these games.

```{r no_winner}
d[d$game == 174 | d$game == 5167, "iswinner"] <- FALSE
```

Game 1776 had a three-way nonzero tie and all players came back as
co-champions.
There are 49 games with a two-way tie after final Jeopardy.
The last tied game was on 2014-10-30.
After that, games that ended in a tie have a single playoff question
to determine a single champion.

After November 23, 2001 the game doubled the dollar amounts for each answer.
A `gamemaxvalue` variable was added to distinguish games before and after the change:

```{r gamemax}
d[["gamemaxvalue"]] <- ifelse(d[["date"]] < as.Date("2001-11-23"),
    "max1000", "max2000")
```

Starting with the September 8, 2003 game winners were no longer limited to
playing only 5 games.
They now return until they lose.
A logical `is5limit` variable was added to distinguish games where
5 game limit (`is5limit TRUE`) was in effect and those without the limit.

```{r is5limit}
d[["is5limit"]] <- d[["date"]] < as.Date("2003-09-08")
```

## Save Clean Data
Save R object of the clean regular games data set.
```{r save_clean_regular}
saveRDS(d, here::here("data/reg_game_player_score.rds"))
```
