---
title: "Cleaning Jeopardy Data"
author: "Stephen Kaluzny"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

## Data

Read the delimited data file.
```{r read_data}
dfile <- "game_player_score.txt"
dpath <- here::here(file.path("data", dfile))
d <- read.table(dpath, header=TRUE, sep="|", quote='"', comment="")
d[["date"]] <- as.Date(d[["date"]])
```

The data has `r nrow(d)` observations and `r ncol(d)` variables.

There are `r length(unique(d$game))` unique games
and `r length(unique(d$show))` unique shows.

The variable names and type.
```{r data_type}
sapply(d, class)
```


## Data Clean Up

The `occupation`, `from`, and `last` variables need work.
We extract them, do the fix up and then put them back into the data.frame.

```{r extract_from_occupation_last}
from2 <- d[["from"]]
occupation2 <- d[["occupation"]]
last2 <- d[["last"]]
```

The scraped data has errors with the `last` name suffix "JR." being included
with `occupation`

```{r clean01}
# Fix occupation and from
# occupation includes "JR.," at start, add to end of last:
indx_jr <- grep("\\bjr\\., ", occupation2, ignore.case=TRUE)
if(length(indx_jr) > 0) {
  occupation2[indx_jr] <- gsub("^JR\\., ", "", occupation2[indx_jr])
  last2[indx_jr] <- paste0(last2[indx_jr] , ", Jr.")
}
```

The word `Originally` gets mixed in with `occupation`.
Just remove it.
The word `From` ends up in `occupation`, remove it.

```{r clean02}
# occupation include "Originally" at the end, remove:
occupation2 <- gsub(" Originally$", "", occupation2)
# No occupation, just the word "From", set to NA:
indxFrom <- which(occupation2 == "From")
if(length(indxFrom) > 0) {
  occupation2[indxFrom] <- NA
}
# No occupation, just the word "Originally", set to NA:
indxOriginally <- which(occupation2 == "Originally")
if(length(indxOriginally) > 0) {
  occupation2[indxOriginally] <- NA
}
# occupation starts with "From" and from is NA, swap:
indxOccHasFromStart <- grep("^From ", occupation2)
indxFromNA <- which(is.na(from2))
indxOccToFrom <- intersect(indxOccHasFromStart, indxFromNA)
if(length(indxOccToFrom) > 0) {
  from2[indxOccToFrom] <- occupation2[indxOccToFrom]
  occupation2[indxOccToFrom] <- NA
}
# occupation ends with "From", remove it:
indxOccHasFromEnd <- grep("\\sFrom$", occupation2)
if(length(indxOccHasFromEnd) > 0) {
  occupation2[indxOccHasFromEnd] <- gsub("\\sFrom$", "",
    occupation2[indxOccHasFromEnd])
}
# Remove "From" from the from variable:
from2 <- gsub("^From ", "", from2)
from2 <- gsub(" From$", "", from2)
```
From previous working with the J Archive data some misspelled
occupation and location values.
Correct with the `fixfrom` function.

```{r clean03}
# Fix known misspelled / problem from values:
fixfrom <- function(x, old, new) {
  indx <- which(x == old)
  if(length(indx) > 0) {
    x[indx] <- new
  }
  x
}
from2 <- fixfrom(from2, "New York City", "New York, New York")
from2 <- fixfrom(from2, "The Bronx", "New York, New York")
from2 <- fixfrom(from2, "West Hartford", "West Hartford, Connecticut")
from2 <- fixfrom(from2, "Lynn Massachusetts", "Lynn, Massachusetts")
from2 <- fixfrom(from2, "San Francisco", "San Francisco, California")
from2 <- fixfrom(from2, "San Jose California", "San Jose, California")
from2 <- fixfrom(from2, "Kansas City", "Kansas City, Missouri")
from2 <- fixfrom(from2, "ST. Louis", "St. Louis, Missouri")
from2 <- fixfrom(from2, "New York State", "New York")
from2 <- fixfrom(from2, "Carmel Valley", "Carmel Valley, California")
from2 <- fixfrom(from2, "Easton Maryland", "Easton, Maryland")
from2 <- fixfrom(from2, "Trinidad And Tobago And Now In New York",
  "New York, New York")
from2 <- fixfrom(from2, "Sandy Mannix", NA)
```

Now add the revised `last2`, `from2` and `occupation2`
to the original data.frame.

```{r update_data.frame}
d[["last2"]] <- last2
d[["from2"]] <- from2
d[["occupation2"]] <- occupation2
```
## Non-Regular Jeopardy Games

Over the years Jeopardy has had a number of special games.
These include an annual Tournament of Champions,
Celebrity Jeopardy, College and Children Tournament.
These games do not follow the standard winner moves on format.
As such, they will be removed.

To find these non-regular games the comments variable is searched
for particular words used in describing these games.

```{r find_non_regular}
indx_tournament <- grep("tournament", d[["comments"]], ignore.case=TRUE)
indx_championship <- grep("championship", d[["comments"]], ignore.case=TRUE)
indx_super_jeopardy <- grep("Super Jeopardy", d[["comments"]], ignore.case=TRUE)
indx_celebrity <- grep("celebrity", d[["comments"]], ignore.case=TRUE)
indx_battle <- grep("\\bbattle\\b", d[["comments"]], ignore.case=TRUE)
indx_kids <- grep("\\bkids week\\b", d[["comments"]], ignore.case=TRUE)
indx_power <- grep("\\bpower players\\b", d[["comments"]], ignore.case=TRUE)
indx_backtoschool <- grep("\\bback to school\\b", d[["comments"]], ignore.case=TRUE)
indx_goat <- grep("\\bgreatest of all time\\b", d[["comments"]], ignore.case=TRUE)
indx_quarterfinal <- grep("quarterfinal", d[["comments"]], ignore.case=TRUE)
indx_semifinal <- grep("semifinal", d[["comments"]], ignore.case=TRUE)
indx_final <- grep("\\bfinal game", d[["comments"]], ignore.case=TRUE)
```

We combine all these sesarch results (indices) and take the union.
```{r indx_not_regular}
indx_not_regular <- sort(unique(c(
  indx_tournament,
  indx_championship,
  indx_super_jeopardy,
  indx_celebrity,
  indx_battle,
  indx_kids,
  indx_power,
  indx_backtoschool,
  indx_goat,
  indx_quarterfinal,
  indx_semifinal,
  indx_final
)))
```

There are some regular games where the words Tournament or Battle were
used in the comments.
We want to keep those games.

```{r keep_games}
# Keep games: 3970 4085 4305 4364,
# Regular game with the word Tournament or Battle in comment:
indx_3970 <- which(d[["game"]] == 3970)
indx_not_regular <- setdiff(indx_not_regular, indx_3970)
indx_4085 <- which(d[["game"]] == 4085)
indx_not_regular <- setdiff(indx_not_regular, indx_4085)
indx_4305 <- which(d[["game"]] == 4305)
indx_not_regular <- setdiff(indx_not_regular, indx_4305)
indx_4364 <- which(d[["game"]] == 4364)
indx_not_regular <- setdiff(indx_not_regular, indx_4364)
```

Now remove all the non-regular games.
Use the regular games data set going forward.

```{r drop_non_regular}
d <- d[-indx_not_regular,]
```

## Missing Values

How many missing values in each variable.

```{r missing}
sapply(d, function(x) sum(is.na(x)))
```

## Data Verification

```{r range_of_date}
range(d$date)
```
